Reconciling agriculture, carbon, and biodiversity in a savanna transformation frontier

### 2016-05-05 paper/main.tex
Full paper revisions: main manuscript, fully revised, with edits and additions to SI methods and results.  New figure created (figures.Rmd), slight edit of loose code on main-analysis.Rmd

Savannas sometimes receive limited respect in global land use assessments and have been a major focus of agricultural expansion, not only in Africa but in places such as the Cerrado of Brazil and northern Thailand \citep{morris_awakening_2009}.  Their carbon contents and biodiversity values may be completely ignored on the grounds that they are not forests \citep{searchinger_high_2015}. Yet, the wetter woodland-savannas and shrublands of Africa have a similar average richness, although not density, for birds and mammals as the world’s wetter tropical forests, and 75\% of these habitats have higher overall vertebrate diversity than 75\% of the rest of the world \citep{searchinger_high_2015}.  Their carbon contents are also substantial, particularly relative to their likely yields, so that their use for food crops is unlikely to result in fewer carbon emissions per ton of crop \citep{searchinger_high_2015}.  Conversion of over 100 million hectares of these lands would release tens of millions of tons of carbon.

The high likelihood of some cropland expansion in sub-Saharan Africa raises the prospect of whether this expansion could be located in ways that allow agricultural potential to be achieved, but for substantially lower loss of carbon and biodiversity than conventional agricultural development pathways.  Analyzing this question must be done at national or sub-national scales, because that is where key decisions are made, and it requires data that is both in resolution and quality to provide confidence for decision-makers.

In this paper we look at Zambia, which is a bellwether for this land use challenge in sub-Saharan Africa. On the one hand, Zambia has great need to boost its food production. Sixty percent of Zambia's population is rural {\citep{the_world_bank_world_2016}, per capita calorie food availability is just 2100 kcal \citep{fao_faostat:_2016}, and child stunting rates are 40\% \citep{haddad_global_2015}.  According to the medium estimate of the United Nations, Zambia’s population is projected to grow 265\% from 16.2 million in 2015 to 43 million in 2050 \citep{united_nations_world_2015}, which will combine with rapid economic growth \citep[$>$6\%,][]{the_world_bank_world_2016} to substantially increase per capita calorie demands \citep{tilman_global_2011}. Zambia has also been expanding its agricultural exports, and has agronomic and situational characteristics that put it in a strong position to continue to do so to other parts of the region.

At the same time, two thirds of Zambia’s 738,400 km$^2$ land area is dominated by woodland savannas, which have sufficiently dense tree cover to be classified as forest under either of the United Nations Framework Convention on Climate Change thresholds ($>$10 or $>$30\% tree cover) for defining this vegetation type \citep{sexton_conservation_2015}.  These ecosystems store substantial amounts of carbon, with 50\% of Zambia having a carbon density $\geq$130 t ha$^{-1}$ (vegetative carbon and soil carbon in the top 1 m), while an additional 40\% exceeds 80 t ha$^{-1}$. Meanwhile, Zambia’s deforestation rate of 2,500--3,000 km$^2$ year$^{-1}$ is one of the highest in the world \citep{vinya_preliminary_2011}.

The biodiversity of these woody savannas is also high.  Most of Zambia is covered by what the World Wildlife Fund has categorized as Central Zambian Miombo Woodland \citep{olson_terrestrial_2001,hogan_central_2004}. It contains mature miombo woodland trees typically 15 to 20 meters tall, around a quarter of which are evergreen, with a broadleaf shrub and grass understory beneath, and is interspersed with wetlands that comprise up to 30\% of the landscape \citep{hogan_central_2004}. In addition to hosting one of the world's three richest mammal assemblages \cite{olson_terrestrial_2001}, the large expanses of relatively intact habitat make these woodlands, and several of Zambia's eight other ecoregions \citep{fund_ecoregions_2011, estes_zambezian_????}, important centers of large mammal populations, including black rhino, African buffalo, and highly migratory or wide-ranging species, such as elephants and major predators \citep{hogan_central_2004}.  Zambia in general has a high bird diversity \citep{searchinger_high_2015}. Plant diversity in much of the country ranges between 3,000-5,000 species, a richness that is exceeded only in tropical forests \citep{kier_global_2005}.  Although there are relatively few endemic bird and mammal species, reptile and amphibian endemism is high, with 19 endemic reptile species and 13 endemic amphibians \citep{fund_ecoregions_2011}.

There is a long history of conservation planning that attempts to target conservation areas or objectives while minimizing economic costs or within a constrained budget of land area for conservation \cite[e.g.][]{koh_spatially_2010, naidoo_integrating_2006,margules_systematic_2000}. Koh and Ghazoul \citep{koh_spatially_2010} provide one example of an effort to examine trade-offs among production, carbon and biodiversity for oil palm development in Indonesia by giving each land use priority equal weighting.  One of the challenges for land use planning in practice is that decisions involve trade-offs among different objectives and values that most people find difficult to weight against each other \emph{ex ante}, because the values are often abstract, making them harder to evaluate without first viewing the consequential trade-offs.  Here we present a new model in which varying weights can be placed on the competing land use objectives. This set of options then allows model users to choose the most appealing combination of weights based on their land use preferences, and their own assessments of what are and are not acceptable costs within those objectives. The objective of this is to assist different stakeholders in finding compromises between competing land use interests while still meeting agricultural production targets. We apply this model to Zambia so that we can explore its strengths and limitations as a decision-making tool, and also to answer a basic but important question: is it possible to achieve a reasonable balance of different interests by reducing the carbon and biodiversity costs of expansion while still focusing on land use ares with high agricultural potential?

Answering this question is becoming increasingly important as savannas and other grassy biomes receive increasing agricultural development pressure, and may become increasingly targeted as an indirect result of the laudable efforts to conserve tropical forests \citep{searchinger_high_2015}.

@@

The model we developed seeks to find optimal land use configurations that satisfy the production targets for multiple crops while minimizing four costs: the total land area required, transportation costs, carbon released from land conversion, and impacts on biodiversity. Although total land area is treated as a cost, this cost is minimized by selecting areas with the highest likely yields. By minimizing total land required as well as transportation costs, the model finds areas that reflect two important indicators of agricultural potential. By minimizing carbon and biodiversity costs, the model finds areas that minimize two important dimensions of environmental impact.  The model chooses final areas depending on the weights that a user assigns to each of these four objectives.

To enable this functionality, and to identify optimal solutions, the model is structured to evaluate each objective in terms of production efficiency, or the ratio of cost per hectare to crop production per hectare (yield) that results from converting land to agriculture.  For land area costs, output is measured in terms of the number of km$^2$ required to meet the crop production target. For the other objectives, the indicators are transportation costs, carbon loss, or biodiversity impacts per ton of crop yield.

We focus our model initially on two crops, maize and soybeans, because of their existing and potential significance for Zambia's agricultural sector, and their likely dominance of in the future growth of agriculture in Africa's broader savanna regions \citep{searchinger_high_2015}.  Maize is the largest crop by area and consumption in Zambia \citep{fao_faostat:_2016}, while soybeans are expanding rapidly as a commodity crop to meet regional and global demands for animal feed and direct human consumption \citep{gasparri_emerging_2015}. Because the model is designed to meet production targets for multiple crops, we apply algorithms for allocating each crop to its highest value cells.

@@

In the second step, we used an empirical model to map the DSSAT-modeled yields onto a 1 km grid, taking advantage of a finer resolution soil map that provides a common frame for estimating both yield potential and soil carbon stocks (see below), but has insufficient detail in several parameters (particularly effective rooting depth) required by DSSAT to provide a sound empirical basis for running the model directly.  We used a generalized additive model \citep{wood_mgcv:_2001} to predict the values of the DSSAT-simulated yields, using best-fitting subsets of the following weather and soil variables: growing season precipitation; growing degree days; mean growing season shortwave solar radiation; percent organic carbon; pH; percent clay content. The soil predictors represented mean values in the top 1 m of soil, and were extracted from a new 1 km global database of soil properties \cite{hengl_soilgrids1km_2014}.  This two step approach allowed us to more soundly capture crop responses to Zambia's climatic variability and potential management practices, and to base potential yields on the same dataset used to estimate carbon costs (see below).

The first two steps resulted in a map of potential yield under high inputs and good management, and assuming no losses due to other limitations such as diseases. These yield values therefore overestimate what farmers are truly likely to achieve. To reflect more realistic estimates of future yields, we therefore rescaled the yield maps to match the mean yields that are projected to be achieved in Zambia by 2050 (4.4 t ha$^{-1}$ for maize and 3.6 t ha$^{-1}$ for soybean).  We projected these yields by using FAO-projections for sub-Saharan Africa as a whole to derive an annual growth factor and applying that factor to the average 2009-2014 yields for each crop reported in Zambia.  Overall, the distribution of yields derived from our crop modeling is most important to targeting land use, while this choice of average yield determines the total quantity of land needed to meet production targets.

@@

As a second measure related to agricultural potential, we estimated the travel time to the nearest major town. Travel time serves here as a proxy for differences in output prices and production costs, such as fertilizer prices and transport costs \cite{stifel_isolation_2008}, and, via the density of road networks, is strongly correlated with the suitability of land for farming \citep{laurance_estimating_2015,estes_using_2014}.  We used a cost-distance analysis drawing on spatial data for roads, rivers and water bodies, urban centers inside and neighboring Zambia, and national boundaries to calculate travel time in hours. Total travel costs for any land use scenario equal the total time needed to transport the tonnage of crops produced in each converted cell to the nearest town, which we divide by 20 to produce "truck hours" on the assumption of a 20-ton truck (although the division  of all cell results by this constant does not alter the mapping outcomes).

@@

To estimate potential carbon costs from cropland conversions, we developed maps of vegetative and soil carbon stocks for Zambia.  Vegetative carbon was calculated from a recent map of above-ground biomass developed by Baccini et al. \citep{baccini_estimated_2012}, from which we estimated below-ground biomass using savanna and miombo woodland-specific root-shoot ratios \cite{mokany_critical_2006, mugasha_allometric_2013}, and then converted the sum to carbon using a ratio of 0.47 \citep{ipcc_land_1996}. We used the same soil carbon dataset used for crop yield prediction together with an accompanying map of bulk density to calculate soil carbon stocks in the top 1 m. The potential carbon loss due to agricultural conversion was calculated as 100\% of vegetative carbon and 25\% of soil carbon \cite{searchinger_high_2015}.

@@

There are many ways of evaluating biodiversity value and potential impacts, including species richness of different taxa in absolute terms or based on level of threat or endemism, and including factors designed to address the significance of contiguity or the impact of different types of development \citep[e.g.][]{newbold_global_2015,searchinger_high_2015, gasparri_emerging_2015,koh_spatially_2010}.  The actual measures selected should reflect both conservation priorities and the quality of available data.  For a fine-scaled, country-level analysis such as this one, the species range map data that form the basis for broader impact assessments \citep[e.g.][]{kehoe_global_2015} have an effective resolution that is too coarse. We therefore developed a composite measure of biodiversity value based on several datasets.

First, we used the potential vegetation maps of East Africa \citep[][]{van_breugel_potential_2011}, a 2010 landcover dataset for Zambia\footnote{downloaded from http://apps.rcmrd.org/landcoverviewer/}, and a shapefile of protected areas to calculate how much of Zambia's 20 unique vegetation types remains after conversion to croplands and settlements, and what proportion of that remainder falls within national parks. From this estimate we calculated a hybrid measure of rarity and threat, wherein we assigned a weight to each vegetation type based on its relative remaining area in proportion to Zambia's extent, log-transformed to account for a highly skewed distribution.  We then re-weighted that weight by the protected fraction of each vegetation type's protected fraction, following \citep{van_breugel_environmental_2015}. We then used the remaining vegetation map to calculate a second measure, the proportion of untransformed vegetation within a 11X11 km neighborhood centered on each 1 km$^2$, providing a measure of vegetation intactness. By adding this measure to the first index and normalizing, we created a biodiversity score (range 0-1 per ha) that gives equal weight to how rare or threatened a habitat is, and how undisturbed its local area.  When the biodiversity score for each converted grid cell converted is summed, it generates a total cost of biodiversity impacts.

As a final modification, we masked out the areas of existing national parks and game management areas to respect the national judgment that they should not be used for expanding farmland.  We retained a third protected category, forest reserves, as these tend to be far less stringently protected, having already had 8.3\% of their area converted to cropland or settlements, but assigned these areas a high biodiversity score.

@@

Likely future demand for maize and soybeans produced in Zambia is difficult to estimate, and our model can easily be run with different production targets.  For this paper, we adopted targets of a four-fold increase for maize for 2050 compared to 2009-2014 average, and a 10-fold increase for soybeans.  These increases reflect the rate of growth in Zambia from 2000-2014, and a mean production trend for soybean growth in southern Africa estimated by Gasparri et al \citep{gasparri_emerging_2015}, with small additional adjustments to to account for the fact that Zambia has one of the highest agricultural potentials in sub-Saharan Africa and has high potential to expand exports.

@@

The amount of land conversion to meet a production target will depend on the growth of yields on existing croplands, which are well short of their current potentials \citep{waddington_getting_2010}.  For the principal scenarios we present in this paper, we assume that yields on current cropland will achieve the average likely yields for Zambia's \emph{potential} farmland.  However, to examine the possible significance of yield gains on existing cropland, we also examined scenarios in which the yield gains on existing cropland 1) only manage to close 50\% of the gap between current and our projected future yields, and 2) close 150\% of the gap.

@@

Weighting different objectives is to some extent based purely on preferences, but it is important that these be expressed in a numerical form and processed so that a quantitatively optimal solution for that set of preferences is identified. To achieve this result, we first express the different objectives as efficiencies (e.g. carbon loss per ton of crop yield, as described above), and then normalize the units of each objective based on the range of its scores. We then multiplying each normalized objective by its corresponding weight (which fall between 0 and 1 and must sum to 1, with values closer to 1 indicating stronger preferences) and then adding the resulting products. The result is a single score that ranks each grid cell according to its ability to meet the production target for the lowest cost, given those weights. The model then selects cells in descending rank order until each the cumulative production of converted cells reaches each crop's corresponding target (see online electronic supplement for details of land allocation between crops).

Although this weighting system treats choices as preferences, our model calculates the actual costs of each objective in absolute terms:  area of land converted (which is increasingly minimized as weight is added to the preference to maximize crop yields), total transportation costs from new croplands, total carbon released from conversions, and total impacts on biodiversity. After reviewing these costs, model users can then adjust the preferences to reflect desirable or politically acceptable trade-offs.

@@

In our main scenario, we assumed that the gap between current and projected future yields would be fully closed on existing cropland, but we also ran the model (with equal weights across all objectives) for cases when only 50\% or 150\% of the gap was closed. In the former case, the total land area required to meet the target increased by 18\% (14,634 to 17,202 km$^2$), transport costs increased 22\% (5.52 X 10$^{-5}$ to 6.61 X 10$^{-5}$ truck hours), carbon losses increased by 19\% (597 to 711 Mt), and the total biodiversity index of converted areas increased by 20\% (5.53 X 10$^{-5}$ to 6.61 X 10$^{-5}$).   In the 150\% gap closure scenario, the impacts were reversed, with impacts declining by almost exactly the same percentage.

@@

The scenario is which the sole preference is to maximize yield (100\% weight) converts land with the most promising yield potential for farmers, which should have economic advantages.  These areas have little direct overlap with the areas picked by scenarios that minimize carbon costs, biodiversity impacts or travel time (Figure 1).  They also have little adjacency, as measured by the average distance between each pixel of one map and its nearest neighbor in the other (Figure 1).

However, the areas selected solely for minimizing carbon, biodiversity, or transportation costs also have little direct overlap, but the map reveals that they have high adjacency, confirmed by direct estimates of distance (Figure 1). All three criteria tend to identify lands along existing road networks and population and agricultural centers, with the carbon minimizing objective converting a larger area of land in the western half of the country. The reasons for this correlation are probably that carbon stocks and biodiversity values are lowest in areas of high human density, because of high levels of wood harvesting for charcoal and fuelwood, and because human uses similarly cause fragmentation that leads to low biodiversity scores.

\begin{figure}[!ht]
  \begin{center}
       \makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{figures/single_priorities3.png}}
    \caption{Areas converted to new maize and soybean croplands and their associated costs when each individual objective (yield maximization, and carbon cost, biodiversity cost, and transport time minimization) is given 100\% weight. Cropland converted under each scenario is shown in the maps, colour-coded according to which objective receives 100\% weight. Large bar charts, colour-coded to the maps, display the absolute value of costs resulting from each scenario. Small bar charts show the percentage of overlap (top) and adjacency (bottom; nearest neighbor distance) between the map directly above it, and each of the other three maps.}
    \label{default}
  \end{center}
\end{figure}

@@

 \begin{figure}[!ht]
  \begin{center}
       \makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{figures/compromise_25pct2.png}}
    \caption{Areas converted to new maize and soybean croplands and their associated costs when when all objectives (yield maximization, and carbon, biodiversity, and transport cost minimization) compromise equally (i.e. receive 25\% weight). Cropland converted under this scenario is denoted by purple areas in the map, with its associated cost impacts shown alongside those from the 5\% compromise scenarios (Figure 3), together with the maximum, minimum, and median costs across all weighting permutations (grey horizontal lines in bar charts).}
    \label{default}
  \end{center}
\end{figure}

>>>>

Savannas sometimes receive limited respect in global land use assessments and have been a major focus of agricultural expansion, not only in Africa but in places such as the Cerrado of Brazil and northern Thailand \citep{morris_awakening_2009}.  Their carbon contents and biodiversity values may be completely ignored on the grounds that they are not forests \citep{searchinger_high_2015}. Yet, the wetter woodland-savannas and shrublands of Africa have a similar average richness, although not density, for birds and mammals as the world's wetter tropical forests, and 75\% of these habitats have higher overall vertebrate diversity than 75\% of the rest of the world \citep{searchinger_high_2015}.  Their carbon contents are also substantial, particularly relative to their likely yields, so that their use for food crops is unlikely to result in fewer carbon emissions per ton of crop \citep{searchinger_high_2015}.  Conversion of over 100 million hectares of these lands would release tens of millions of tons of carbon \citep{searchinger_high_2015}.

The high likelihood of some cropland expansion in sub-Saharan Africa raises the prospect of whether this expansion could be located in ways that meet agricultural production goals, but for substantially lower loss of carbon and biodiversity than conventional agricultural development pathways.  Analyzing this question must be done at national or sub-national scales, because that is where key decisions are made, and accurate results require data that are high in both resolution and quality.

In this paper we look at Zambia, which is a bellwether for this land use challenge in sub-Saharan Africa. On the one hand, Zambia has great need to boost its food production. Sixty percent of Zambia's population is rural {\citep{the_world_bank_world_2016}, per capita calorie food availability is just 2100 kcal \citep{fao_faostat_2016}, and child stunting rates are 40\% \citep{haddad_global_2015}.  According to the medium estimate of the United Nations, Zambia's population is projected to grow 265\% from 16.2 million in 2015 to 43 million in 2050 \citep{united_nations_world_2015}, which will combine with rapid economic growth \citep[$>$6\%,][]{the_world_bank_world_2016} to substantially increase per capita calorie demands \citep{tilman_global_2011}. Zambia has also been expanding its agricultural exports, and has agronomic and situational characteristics that put it in a strong position to continue to do so to other parts of the region.

At the same time, two thirds of Zambia's 738,400 km$^2$ land area is dominated by woodland savannas, which have at sufficiently dense tree cover to be classified as forest under either of the definitions ($>$10 or $>$30\% tree cover) used by the United Nations Framework Convention on Climate Change \citep{sexton_conservation_2015}.  These ecosystems store substantial amounts of carbon, with 50\% of Zambia having a carbon density of at least 130 t ha$^{-1}$ (vegetative carbon and soil carbon in the top 1 m), while an additional 40\% exceeds 80 t ha$^{-1}$. Meanwhile, Zambia's deforestation rate of 2,500--3,000 km$^2$ year$^{-1}$ is one of the highest in the world \citep{vinya_preliminary_2011}.

The biodiversity of these woody savannas is also high. Most of Zambia is covered by what the World Wildlife Fund has categorized as the Central Zambian Miombo Woodland ecoregion \citep{olson_terrestrial_2001,hogan_central_2004}, which contains some 3800 plant species, making it the 17th richest ecoregion in the world in floral diversity (out of 867) \citep{kier_global_2005}. Vegetation is characterized by a mix of tall, frequently evergreen, trees (15-20 m), broadleaf shrubs, and grasses, and is heavily interspersed with wetlands \citep{hogan_central_2004}.  The large expanses of relatively intact habitat give this woodland the third highest richness for mammal species of the worlds ecoregions  \citep{olson_terrestrial_2001,fund_ecoregions_2011, estes_zambezian_????}.  It supports rare species including the  black rhino,  major predators, and highly migratory, wide-ranging species, such as elephants \citep{hogan_central_2004}.  Zambia overall also has a high bird diversity, with 753 recorded species \citep{dowsett_birds_2008}. Although few bird and mammal species are endemic, reptile and amphibian endemism is high, with 19 endemic reptile species and 13 endemic amphibians \citep{fund_ecoregions_2011}.

There is a long history of using models to achieve conservation objectives in the face of cost and other constraints \cite[e.g.][]{naidoo_integrating_2006,margules_systematic_2000}. MARXAN, perhaps the most widely used of these, is a decision-support program that is designed to efficiently select conservation area networks \citep{ball_marxan_2009,watts_marxan_2009}. Within the more general land use planning field, an even broader range of models exists for finding optimal spatial configurations that satisfy multiple land use objectives (see Cao et al. \citep{cao_spatial_2011} for an example and review of alternatives). In the present case, the need is not to select protected areas but to identify areas for cropland expansion that meet the objectives of agricultural development and environmental conservation. Koh and Ghazoul \citep{koh_spatially_2010} developed one model for examining trade-offs among oil palm development, rice production, and carbon and biodiversity production in Indonesia, and a related approach was used to identify sustainable agricultural intensification solutions in southern Tanzania \citep{nijbroek_regional_2016}. In both examples, the model assigned equal weight to each land use objective in the compromise scenarios that were analyzed. However, one of the great challenges for land use planning is that different objectives can be difficult to weigh against each other \emph{ex ante}, because the values are often abstract, making them harder to evaluate without first viewing the consequential trade-offs \citep{cao_spatial_2011,nijbroek_regional_2016}. Governments and other stakeholders also tend to have varying, and often conflicting, priorities, and their willingness to compromise may depend on the degree to which their priorities are impacted.

To overcome these problems, we developed a new, publicly available model that allows varying levels of compromise between competing agricultural and environmental interests to be evaluated while still meeting agricultural production targets. The model adopts different methods than the two aforementioned models \citep{koh_spatially_2010,nijbroek_regional_2016} to estimate agricultural potential and conservation objectives, and can allocate land for multiple crops simultaneously. More fundamentally, our model is based on linear programming, which allows users to place varying weights on the different land use objectives \citep{cao_spatial_2011}. This feature allows different stakeholders to work together to choose the most appealing combination of weights based on their land use preferences, and their own assessments of what are and are not acceptable costs within those objectives.

We apply this model to Zambia so that we can explore its strengths and limitations as a decision-making tool, and also to answer a basic but important question: is it possible to achieve a reasonable balance of different interests by reducing the carbon and biodiversity costs of expansion while still focusing on land use areas with high agricultural potential?  Answering this question is becoming increasingly important as savannas and other grassy biomes receive increasing agricultural development pressure, and may become increasingly targeted as an indirect result of the laudable efforts to conserve tropical forests \citep{searchinger_high_2015}.

@@

The model we developed seeks to find optimal land use configurations that satisfy the production targets for multiple crops while minimizing four costs: the total land area required, transportation costs, carbon released from land conversion, and impacts on biodiversity. Although total land area is treated as a cost, minimizing this cost identifies lands with the highest likely yields. By minimizing total land required as well as transportation costs, the model therefore finds areas that reflect two important indicators of agricultural potential. By minimizing carbon and biodiversity costs, the model also finds areas that minimize two important dimensions of environmental impact.  The model chooses final areas depending on the weights that a user assigns to each of these four objectives.

To enable this functionality, and to identify optimal solutions, the model is structured to evaluate each objective in terms of production efficiency, or the ratio of each cost per hectare to crop production per hectare (yield) that results from converting land to agriculture.  For land area, cost is measured as the number of km$^2$ required to meet the crop production target. For the other objectives, the costs indicators are transportation time, carbon loss, or biodiversity impacts.

We focus our model initially on two crops, maize and soybeans, because of their existing and potential significance for Zambia's agricultural sector, and their likely dominance of in the future growth of agriculture in Africa's broader savanna regions \citep{searchinger_high_2015}.  Maize is the largest crop by area and consumption in Zambia \citep{fao_faostat_2016}, while soybeans are expanding rapidly as a commodity crop to meet regional and global demands for animal feed and direct human consumption \citep{gasparri_emerging_2015}. Because the model is designed to meet production targets for multiple crops, we apply algorithms for allocating each crop to its highest value cells, given the particular combination of objectives being assessed.

@@

In the second step, we used an empirical model to map the DSSAT-modeled yields onto a 1 km grid, taking advantage of a finer resolution soil map that provides a common frame for estimating both yield potential and soil carbon stocks (see below), but has insufficient detail in several parameters (particularly effective rooting depth) required by DSSAT to provide a sound empirical basis for running the model directly.  We used a generalized additive model \citep{wood_mgcv_2001} to predict the values of the DSSAT-simulated yields, using best-fitting subsets of the following weather and soil variables: growing season precipitation; growing degree days; mean growing season shortwave solar radiation; percent organic carbon; pH; percent clay content. The soil predictors represented mean values in the top 1 m of soil, and were extracted from a new 1 km global database of soil properties \cite{hengl_soilgrids1km_2014}.  This two step approach allowed us to more soundly capture crop responses to Zambia's climatic variability and potential management practices, and to base potential yields on the same dataset used to estimate carbon costs (see below).

The first two steps resulted in a map of potential yield under high inputs and good management, and assuming no losses due to other limitations such as diseases. These yield values therefore overestimate what farmers are truly likely to achieve. To reflect more realistic estimates of future yields, we therefore rescaled the yield maps to match FAO-projected Zambian yields in 2050 (4.4 t ha$^{-1}$ for maize and 3.6 t ha$^{-1}$ for soybean).  We used the FAO-projections for sub-Saharan Africa as a whole to derive an annual growth factor and applied that factor to the average 2009-2014 yields for each crop in Zambia. Overall, the distribution of yields derived from our crop modeling is most important to targeting land use, while this choice of average yield determines the total quantity of land needed to meet production targets.

@@

As a second measure related to agricultural potential, we estimated the travel time to the nearest major market town. Travel time serves here as a proxy for differences in output prices and production costs, such as fertilizer and transport costs \cite{stifel_isolation_2008}, and, via the density of road networks, is strongly correlated with the suitability of land for farming \citep{laurance_estimating_2015,estes_using_2014}. For this analysis, we defined market towns as the administrative capitals of Zambia's districts, as well as any town outside of Zambia having a population of $\geq$10,000 within a region bounded by 19.5$^{\circ}$ to 34.5$^{\circ}$ longitude and -19.5$^{\circ}$ to -8$^{\circ}$ latitude. These neighboring towns were included because they may be more easily accessible to Zambians living in border regions than the nearest district capital.

We calculated travel time in hours using cost-distance analysis drawing on several spatial datasets. Vector data for Zambian roads obtained from the Government of Zambia (GOZ) and for the broader region from OpenStreetMap\footnote{http://openstreetmap.org} were merged to create a single dataset of regional roads that distinguished between trunk, primary, secondary, and tertiary roads. For market towns, we obtained district capital towns locations from the GOZ and for towns outside Zambia we used data from the Gridded Population of the World Version 3 \citep{ciesin_gridded_2005} database. For waterbodies, we used the USGS Hydrosheds database \citep{lehner_new_2008} to indicate the position and size of rivers, based on their contributing area, and used a lakes vector provided by the GOZ. The roads, rivers, and lakes data, as well as Zambia's border, were converted to 1 km grids wherein each feature was assigned a travel time value based on how long it would take to traverse 1 km. The resulting grids were merged into a single ``friction'' surface, which was used to compute the number of travel hours along the fastest route to the nearest market town (see online electronic supplement for further details).

Total travel costs for any land use scenario equal the total time needed to transport the tonnage of crops produced in each converted cell to the nearest town, We divided these costs by 20 to produce "truck hours" on the assumption of a 20-ton truck. Doing so produces a more intuitive unit, but does not alter mapping outcomes.

@@

To estimate potential carbon costs from cropland conversions, we developed maps of vegetative and soil carbon stocks for Zambia.  Vegetative carbon was calculated from a recent map of above-ground biomass developed by Baccini et al. \citep{baccini_estimated_2012}. We estimated below-ground biomass using savanna and miombo woodland-specific root-shoot ratios \cite{mokany_critical_2006, mugasha_allometric_2013}, and then converted the total biomass to carbon using a ratio of 0.47 \citep{ipcc_land_1996}. The same soil carbon dataset used for crop yield prediction, together with an accompanying map of bulk density, was used to calculate soil carbon stocks in the top 1 m. We calculated the potential carbon loss due to agricultural conversion was calculated as 100\% of vegetative carbon and 25\% of soil carbon \cite{searchinger_high_2015}.

@@

There are many ways of evaluating biodiversity value and potential impacts, including species richness of different taxa in absolute terms or based on level of threat or endemism \citep[e.g.][]{searchinger_high_2015, gasparri_emerging_2015}. Other possibilities include factors designed to address the significance of contiguity or the impact of different types of development on diversity \citep[e.g.][]{newbold_global_2015,koh_spatially_2010}.  The actual measures selected should reflect both conservation priorities and the quality of available data.  For a fine-scaled, country-level analysis such as this one, the species range map data that form the basis for many broader impact assessments \citep[e.g.][]{kehoe_global_2015} have an effective resolution that is too coarse. We therefore developed a composite measure of biodiversity value based on several datasets.

First, we used the potential vegetation maps of East Africa \citep[][]{van_breugel_potential_2011}, a 2010 landcover dataset for Zambia\footnote{downloaded from http://apps.rcmrd.org/landcoverviewer/}, and a shapefile of protected areas to calculate how much of Zambia's 20 unique vegetation types remained after conversion to croplands and settlements as of 2010, and what proportion of that remainder fell within national parks. We used these values to calculate an index of rarity and threat. We calculated rarity as the proportion of Zambia occupied by each vegetation type's remaining area, which we then log-transformed to account for the highly skewed distribution. For threat, we calculated the proportion of each vegetation type falling within protected areas, and then multiplied the threat and rarity values to create the index \citep{van_breugel_environmental_2015}. We then used the remaining vegetation map to calculate a second measure, the proportion of untransformed vegetation within a 11x11 km neighborhood centered on each 1 km$^2$. By adding this measure to the first index and normalizing, we created a biodiversity score with a range of 0-1 per ha, which gives equal weight to a habitat based on a) how rare or threatened and b) how undisturbed it is.  Summing the biodiversity score for each converted grid cell generates a total cost of biodiversity impacts for each scenario.

As a final modification, we masked out the areas of existing national parks and game management areas to respect the national judgment that they should not be used for expanding farmland.  We retained a third protected category, forest reserves, as these tend to be far less stringently protected, having already lost 8.3\% of their area to cropland or settlements, but assigned these areas a high biodiversity score (0.927, the inverse of their converted proportion).

@@

Likely future demand for maize and soybeans produced in Zambia is difficult to estimate, and our model can easily be run with different production targets.  For this paper, we adopted targets of a four-fold increase for maize for 2050 compared to 2009-2014 average, and a 10-fold increase for soybeans.  These increases reflect the rate of growth in Zambia from 2000-2014, and a mean production trend for soybean growth in southern Africa estimated by Gasparri et al \citep{gasparri_emerging_2015}, with small additional adjustments to account for the fact that Zambia has one of the highest agricultural potentials in sub-Saharan Africa and has high potential to expand exports.

@@

The amount of land conversion to meet a production target will depend on the growth of yields on existing croplands, which are well short of their current potentials \citep{waddington_getting_2010}.  For the principal scenarios we present in this paper, we assume that yields on current cropland will achieve the average likely yields for Zambia's \emph{potential} farmland.  However, to examine the possible significance of yield gains on existing cropland, we also examined scenarios in which the yield gains on existing cropland 1) only manage to close 50\% of the gap between current and our projected future yields, and 2) exceed the gap by 50\% (i.e. achieve 150\% of the expected yield gain).

@@

Weighting different objectives to some extent depends purely on preferences, but it is important that these preferences be expressed in a numerical form that reasonably reflects them so that a quantitative solution can be computed. To achieve this result, we first express the different objectives as efficiencies, e.g. carbon loss per ton of crop yield, and then normalize the units of each objective based on the range of its scores. Weights representing the percentage of importance are placed on each objective, expressed as a decimal between 0 and 1, with values closer to 1 indicating stronger preferences, and with all weights summing to 1. The final score is the sum of the resulting products for each objective for each grid cell, which is used to rank each grid cell according to its ability to meet the production target most for the lowest total cost, given those weights. The model then selects cells in descending rank order until the cumulative production of converted cells reaches each crop's corresponding production target.

Although this weighting system treats choices as relative preferences, our model calculates the actual costs of each objective in absolute terms:  total area converted, total transportation costs from new croplands, total carbon released from land conversions, and total impacts on biodiversity. After reviewing these costs, model users can then adjust the preferences to reflect desirable or politically acceptable trade-offs.

@@

In our main scenario, we assumed that the gap between current and projected future yields would be fully closed on existing cropland, but we also ran the model for cases when only 50\% or 150\% of the gap was closed (with equal weights across all objectives). In the former case, the total land area required to meet the target increased by 18\% (14,634 to 17,202 km$^2$), transport costs increased 22\% (5.52 x 10$^{-5}$ to 6.61 x 10$^{-5}$ truck hours), carbon losses increased by 19\% (597 to 711 Mt), and the total biodiversity index of converted areas increased by 20\% (5.53 x 10$^{-5}$ to 6.61 x 10$^{-5}$).   In the 150\% gap closure scenario, the impacts were reversed, with impacts declining by almost exactly the same percentage.

@@

The scenario designed solely to maximize yield (100\% weight) converts land with the most promising yield potential for farmers, which should have economic advantages.  These areas have little direct overlap with the areas picked by scenarios that minimize carbon costs, biodiversity impacts or travel time (Figure 1).  They also have little adjacency, as measured by the average distance between each pixel of one map and its nearest neighbor in the other (Figure 1).

The areas selected solely for minimizing carbon, biodiversity, or transportation costs also have little direct overlap, but the map reveals that they have high adjacency, confirmed by direct estimates of distance (Figure 1). All three criteria tend to identify lands along existing road networks and population and agricultural centers, with the carbon minimizing objective converting a larger area of land in the western half of the country. The reasons for this correlation are probably that carbon stocks and biodiversity values are lowest in areas of high human densities, which have correspondingly high levels of wood harvesting for charcoal and fuelwood, and higher levels of habitat loss and fragmentation leading to lower biodiversity scores.

\begin{figure}[!ht]
  \begin{center}
       \makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{figures/figure1.pdf}}
    \caption{Areas converted to new maize and soybean croplands and their associated costs when each individual objective (yield maximization, and carbon cost, biodiversity cost, and transport time minimization) is given 100\% weight. Cropland converted under each scenario is shown in the maps, colour-coded according to which objective receives 100\% weight. Large bar charts, colour-coded to the maps, display the absolute value of costs resulting from each scenario. Small bar charts show the percentage of overlap (top) and adjacency (bottom; nearest neighbor distance) between the map directly above it, and each of the other three maps.}
    \label{default}
  \end{center}
\end{figure}

@@

 \begin{figure}[!ht]
  \begin{center}
       \makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{figures/figure5.pdf}}
    \caption{Areas converted to new maize and soybean croplands and their associated costs when when all objectives (yield maximization, and carbon, biodiversity, and transport cost minimization) compromise equally (i.e. receive 25\% weight). Cropland converted under this scenario is denoted by purple areas in the map, with its associated cost impacts shown alongside those from the 5\% compromise scenarios (Figure 4), together with the maximum, minimum, and median costs across all weighting permutations (grey horizontal lines in bar charts).}
    \label{default}
  \end{center}
\end{figure}
